### 2015-07-03 thesis.tex
commit for merge

% Template file for a standard thesis
\documentclass[11pt]{isuthesis}
%---------------------------------------------------
% \usepackage{color}
% \usepackage[dvipsnames,svgnames,table]{xcolor}
\usepackage{wrapfig,float}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage{ulem}
\usepackage[section]{placeins}
\usepackage{multirow}
\usepackage{bbm}
\usepackage{rotating}
\usepackage[referable]{threeparttablex}
\usepackage{footnote} % table footnotes
\usepackage{multicol}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{Figure/}}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{url}
\usepackage{bbm}
\usepackage[titletoc]{appendix}

@@

\begin{document}
\DeclareGraphicsExtensions{.jpg,.pdf,.mps,.png}
%-------------------------------------------------------------------------------
\include{Preface/titlepage}
%-------------------------------------------------------------------------------
% Optional thesis dedication
% \include{Preface/dedication}
%-------------------------------------------------------------------------------
% Table of Contents, List of Tables and List of Figures
\pdfbookmark[1]{TABLE OF CONTENTS}{table}
\tableofcontents
\addtocontents{toc}{\def\protect\@chapapp{}} \cleardoublepage \phantomsection
\addcontentsline{toc}{chapter}{LIST OF TABLES}
\listoftables
\cleardoublepage \phantomsection \addcontentsline{toc}{chapter}{LIST OF FIGURES}
\listoffigures
% Comment out the next line if NOT using chaptertitle
\addtocontents{toc}{\def\protect\@chapapp{CHAPTER\ }}
%-------------------------------------------------------------------------------
%Optional Dedication
\cleardoublepage \phantomsection
\include{Preface/dedication}
%-------------------------------------------------------------------------------
%Optional Acknowledgements
\cleardoublepage \phantomsection
\include{Preface/acknowl}
%-------------------------------------------------------------------------------
%Optional thesis abstract
% \cleardoublepage \phantomsection
% \include{Preface/abstract}
%-------------------------------------------------------------------------------
\newpage
\pagenumbering{arabic}
%-------------------------------------------------------------------------------

@@

>>>>

% Template file for a standard thesis
\documentclass[11pt]{isuthesis}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
%---------------------------------------------------
% \usepackage{color}
% \usepackage[dvipsnames,svgnames,table]{xcolor}
\usepackage{wrapfig,float}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage{ulem}
\usepackage[section]{placeins}
\usepackage{multirow}
\usepackage{bbm}
\usepackage{rotating}
\usepackage[referable]{threeparttablex}
\usepackage{footnote} % table footnotes
\usepackage{multicol}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{fullpage}
\usepackage{graphicx,float,wrapfig,subfig,tabularx,ulem}
\graphicspath{{Figure/}}
%\usepackage{csquotes}
\usepackage{color}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{url}
\usepackage{bbm}
\usepackage[titletoc]{appendix}

@@

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\DeclareGraphicsExtensions{.jpg,.pdf,.mps,.png}
%-------------------------------------------------------------------------------
\include{Preface/titlepage}
%-------------------------------------------------------------------------------
% Optional thesis dedication
% \include{Preface/dedication}
%-------------------------------------------------------------------------------
% Table of Contents, List of Tables and List of Figures
\pdfbookmark[1]{TABLE OF CONTENTS}{table}
\tableofcontents
\addtocontents{toc}{\def\protect\@chapapp{}} \cleardoublepage \phantomsection
\addcontentsline{toc}{chapter}{LIST OF TABLES}
\listoftables
\cleardoublepage \phantomsection \addcontentsline{toc}{chapter}{LIST OF FIGURES}
\listoffigures
% Comment out the next line if NOT using chaptertitle
\addtocontents{toc}{\def\protect\@chapapp{CHAPTER\ }}
%-------------------------------------------------------------------------------
%Optional Dedication
\cleardoublepage \phantomsection
\include{Preface/dedication}
%-------------------------------------------------------------------------------
%Optional Acknowledgements
\cleardoublepage \phantomsection
\include{Preface/acknowl}
%-------------------------------------------------------------------------------
%Optional thesis abstract
% \cleardoublepage \phantomsection
% \include{Preface/abstract}
%-------------------------------------------------------------------------------
\newpage
\pagenumbering{arabic}
%-------------------------------------------------------------------------------

@@

\chapter{COMPARISON OF LEARNING OUTCOMES FOR SIMULATION-BASED AND TRADITIONAL INFERENCE CURRICULA IN A DESIGNED EDUCATIONAL EXPERIMENT}
\label{CurriculumStudy}

\begin{center}
\textbf{Status}: Submitted for Review to \textit{Technology Innovations in Statistics Education} (TISE)\\
\end{center}

\begin{center}
\textbf{Authors}\\
Karsten Maurer, Iowa State University, Primary Author\\
Dennis Lock, Iowa State University
\end{center}

\begin{center}
\textbf{Abstract}\\
\end{center}

Conducting inference is a cornerstone upon which the practice of statistics is based. As such, a large portion of most introductory statistics courses is focused on teaching the fundamentals of statistical inference. The goal of this study is to make a formal comparison of learning outcomes under the traditional and simulation-based inference curricula. A randomized experiment was conducted to administer the two curricula to students in an introductory statistics course.  The results indicate that students receiving the simulation-based curriculum have significantly higher learning outcomes for confidence interval related topics. While the results are not comprehensive in assessing the effect on all facets of learning, they indicate that learning outcomes for core concepts of statistical inference can be significantly improved with the simulation-based approach.

%---------------------------------------------------------------------------
\section{Introduction}
\label{intro}

Conducting inference is a cornerstone upon which the practice of statistics is based. As such, a large portion of most introductory statistics courses is focused on teaching the fundamentals of statistical inference. In recent years the approach by which to teach inference in introductory statistics courses has been the topic of growing discussion.  The traditional approach to inference curriculum is focused on distributional theory-based methodology, often characterized by use of distributional assumptions, formulas and tables. A modern alternative is a simulation-based approach to the inference curriculum. The simulation-based approach utilizes tactile and computational simulation to run inferential techniques such as bootstrapping for confidence intervals and simulation-based hypothesis testing. Many proponents of the simulation-based inference curriculum argue that this allows students to be exposed to the core concepts of the inference without first requiring the understanding of theoretical probability distributions.

The focus of the following study is to make a formal comparison of learning outcomes under the traditional and simulation-based inference curricula.  The learning outcomes for concepts surrounding inference with confidence intervals and hypothesis testing are of primary interest.  A randomized experiment was conducted to administer the two curricula to students in an introductory statistics course. The experimental design allows for causal inference to be drawn about the effect of curriculum type on the learning outcomes. The results indicate that significant improvement in learning outcomes for confidence interval related topics are achieved using the simulation-based teaching methods.
%---------------------------------------------------------------------------
\section{Literature Review}
\label{litreview}

With the goal to make proper comparison of traditional versus simulation-based curricula for introductory statistical courses, we must first view where each approach stands within the constant evolution of statistics education. Using the term ``traditional'' to describe the current standard for introductory statistics course curriculum is relative to only the last two decades. Moore chronicled the reform movement of statistic education of the 1980's and 1990's as a period of drastic change in the introductory statistics classroom.  The curriculum expanded greatly from a course dominated by theory-based inference methodology to the inclusion of the topics of data exploration, data production, model diagnostics and simulation.  The content change indicated a shifting emphasis toward conceptual understanding and applied statistics.  Moore also stated, ``(w)hat is striking about the current reform movement is not only its momentum but the fact that it centers on pedagogy as much as content'' \citep{Moore1997}. The pedagogical push toward active learning was combined with content change and the increasing use of technology to form what may be referred to now as the traditional introductory statistics curriculum.

The tenets of the statistics education reform movement were formalized in the Guidelines for Assessment and Instruction in Statistics Education (GAISE) reports for pre-K-12 \citep{GAISEk12} and introductory college courses \citep{GAISEcollege}. Six recommendations were made in the executive summary of the GAISE college report: emphasize statistical literacy and thinking, use real data, stress conceptual over procedural understanding, foster active learning, use technology for both learning and analysis, and use assessment as part of the learning process.  In the past decade these principles have been widely adopted in statistics education with a noteworthy increase in technological integration.  Technology in the statistics classroom now regularly takes the form of applets, graphing calculators, multimedia materials, and educational, analytical and graphical software (\citealt{chance2007}; \citealt{Rubin2007}).  Technological proliferation in the statistics classroom came as a result of technologically receptive statistics educators taking advantage of computation that has become cheaper and more accessible.  A large survey of introductory statistics instructors found that 76\% of the instructors usually or always require students to use a computer program to explore and analyze data, and 90\% of the instructors report a high level of comfort using computer applications to teach introductory statistics \citep{Hassad2013}.

Amidst the drastic increase in the use of technology in introductory statistics education there has been a growing group of educators who believe that the curriculum reform has stopped short of the possibilities that computation can provide.  Cobb argues that statistics education has done well to adopt technology to displace tedious calculation but has not effectively changed the approach to teaching inference.  Cobb strongly articulates a call for statistics instructors to use simulation-based methods for teaching inference to replace the traditional approach to inference using theory-based methodology.  He states, ``(o)ur curriculum is needlessly complicated because we put the normal distribution... at the center of our curriculum, instead of the core logic of inference at the center'' \citep{Cobb2007}.  If we view the introductory statistics course as a constrained optimization problem with statistical literacy and conceptual understanding of inference as the items to maximize, then removing the burden of learning the normal distribution will present the opportunity for more time spent learning core concepts \citep{Carver2011}.   In recent years, curricula for using a simulation-based approach to inference have been developed by a number of groups of statistics educators (\citealt{ISI}; \citealt{Lock5}; \citealt{CATALST};  \citealt{Carver2011}).

There has been research done on the efficacy of simulation-based inference curricula; however, due to the recency of the curricula development most of this preliminary research has been observational.  Budgett, Pfannkuch, Regan \& Wild conduct a case study on a small group of students receiving a simulation-based curriculum and found significant learning gains using pre and post testing based on the Comprehensive Assessment of Outcomes in a First Statistics Course (CAOS).  This study does not however attempt to make a comparison between the simulation-based approach and traditional approach to teaching inference \citep{Budgett2013}. Another pair of studies make comparisons on both learning outcomes and learning retention between the two types of curricula. Tintle, VanderStoep, Holmes, Quisenberry and Swanson found weak evidence for an overall improvement in learning outcomes and significant improvements within the topic of hypothesis testing for the cohort of students receiving the simulation-based curriculum, but the lack of random assignment of student to cohort obstructs the ability to draw any causal conclusions \citep{Tintle2011}.  Tintle, Topliff, VanderStoep, Holmes and Swanson then found significant evidence for improvements to learning outcome retention after four months for students receiving the simulation-based inference curriculum, but again self-selection of students to cohort prevents establishing a causal link \citep{Tintle2012}.

The preliminary research shows promising results for the simulation-based approach to teaching statistical inference.  A more rigorous experimental approach to comparing the traditional and simulation-based curricula has been taken in this study in order to establish a causal effect of curriculum on learning outcomes. Section 3 explains the structure and methodology implemented in the educational experiment and the measurement of student learning.  Section 4 details the model based approach for assessing the effect of curriculum on specific learning outcomes.  Lastly, we discuss the study findings and explore the implications for designing future introductory statistics curricula.
%---------------------------------------------------------------------------
\section{Methodology}
\label{methods}

The subjects for this study were students enrolled in two sections of the Introduction to Statistics, Stat 104, course at Iowa State University in the spring semester of 2014. Stat 104 is an introductory statistics course tailored for students in the agricultural and biological sciences. Of the 112 students to complete the course, 101 students consented to the release of their course data for the purposes of this study.   The students who did not consent were treated identically to those who consented, but their data was omitted from the analysis that follows.  Students from both sections were randomly assigned to one of the two inference curriculum treatments, creating cohorts A,C and B,D, respectively. Cohorts A and B were exposed to the simulation-based curriculum; while the cohorts C and D were exposed to the traditional curriculum. Student cohorts were the basic units to which room assignments, instruction and curriculum treatments were applied.

The course was administered by the authors in a co-teaching setting for students from all cohorts. The course schedule involved two hours of lecture and two hours of lab per week.  The co-teaching strategy was employed as an intentional attribute of the experimental design. The following subsections will detail the curriculum outline for each cohort of students, the experimental design for administering the curricula using the strengths of the co-teaching setup and the data collected for analysis.
%-----------------------------------------------------
\subsection{Curricula Structures}
\label{curric}

To compare the learning outcomes for students receiving the traditional and simulation-based inference curricula we first needed to prepare a curriculum for each approach.  Both curricula needed to satisfy the course guidelines set by the Department of Statistics at Iowa State University, covering the following topics: univariate and bivariate descriptive statistics, linear regression, experimental design, basic probability rules, the binomial distribution, the normal distribution, sampling distributions, and inference on means and proportions. Each curriculum was composed of lecture, corresponding lecture notes, weekly lab assignments designed for groups of four to five students, weekly homework assignments, a midterm exam and a cumulative final exam.  The curricula materials for the course did not require the use of a textbook, however specific textbooks that roughly follow the structure of each curriculum were recommended as supplementary study materials (\citealt{AgrestiFranklin}; \citealt{Lock5}).

Figure~\ref{fig:CurricSched} outlines how these topics were structured within a weekly schedule for the sixteen week semester for each curriculum.  Note that students from all cohorts were exposed to an identical curriculum for all non-inference related topics in the course. This includes identical lecture, course notes, homework assignments, lab assignments and midterm exam during the first half of the semester.

Starting at week 9 the curricula diverge into their respective approaches to inference.  Cohorts A and B began the simulation-based inference curriculum in week 9 by first learning the concepts of sampling distributions then used computer simulation and sampling variability as a basis for exploring inference using bootstrap confidence intervals and simulation-based tests.  To be specific, confidence intervals were constructed by estimating the standard error using the standard deviation of the bootstrap distribution, not through percentile-based bootstrap methods.  Lectures, homework and labs for these cohorts utilized the StatKey software package \citep{Lock5} to conduct the simulation-based inference. The simulation-based curriculum then covered normal distributions and how they could be used to conduct inference on means and proportion.  While many advocates for simulation-based methods may argue that the normal distribution should be pushed to a second course in statistics, course guidelines required that all students of this introductory statistics course be taught theory-based inference methodology.

\begin{figure}[hbtp]
\centering
\includegraphics[keepaspectratio=true, width=1\textwidth]{CurriculumStudy/CurriculaPaths2.pdf}
\caption{\label{fig:CurricSched} Curricula Schedules}
\end{figure}

Cohorts C and D progressed through the traditional approach by first learning the normal distribution and use of the normal tables. They were then introduced to applications of the normal approximation within inference.  The traditional curriculum utilized simulation to display concepts, but only to the extent of demonstrating that sampling distributions can be approximated by normal distributions under certain conditions.

During the second half of the semester the lectures, course note, homework and lab assignments differed between the two curricula.  However, homework and lab assignments were kept similar when they covered similar topics.  For example, all cohorts covered the topic of sampling distributions so the lab assignments were nearly identical between the two groups with the exception of a question pertaining to the normal approximation included for the traditional cohorts.  By the end of the semester all cohorts covered how to conduct inference using normal theory; however cohorts A and B additionally learned the core concepts of inference using simulation-based methods prior to learning traditional theory-based inference methods.

%-----------------------------------------------------
\subsection{Experimental Design}
\label{design}

The logistics of administering a course with two distinct curricula and four cohorts of students required a well-structured design and creative scheduling on several fronts.  The primary objectives for the experimental design were to eliminate differences in non-inference related curriculum administration to the extent possible, remove the confounding instructor effect on each curriculum and to mitigate the effect of unknown lurking variables through random assignment of students to curricula.

Students were randomly assigned to cohorts during the first week of the course.  Of the 101 students who completed the course and consented to the release of their data there were 50 students in the traditional treatment group and 51 students in the simulation-based treatment group. It is also worthwhile to note that of the 4 students to drop the course, all did so prior to week 9; thus, we can safely assume that the inference curriculum treatment did not play a role in the drop. All students who began the inference curricula completed the course.

Students were exposed to identical lecture and lab instruction for the first half of the semester and then diverge into two separate lecture and lab settings for the second half of the semester. This was done to make the experience as similar as possible such that both treatment groups would have the same exposure to terminology and ideas leading up to the inference topics. We could not reassign students to lecture and lab times different than the times for which they enrolled, which meant the logistics of the design required preemptive room scheduling and course time scheduling preparations.  By working with the department chair and course coordinator before students enrolled into sections, we were able to schedule two sections of the course to have identical lecture times but separate lab times. Special room scheduling was required because all students needed to attend the same lecture and lab rooms for the first half of the semester then split into separate lecture and lab rooms after the midterm.  This room and course time scheduling allowed for students to be divided into cohorts and attend the lecture or lab specific to their curriculum.  The lecture and lab room schedules for each cohort are displayed in Figure~\ref{fig:InstSched}.

\begin{figure}[hbtp]
\centering
\includegraphics[keepaspectratio=true, width=1\textwidth]{CurriculumStudy/LectureInstructionWithRooms.pdf}\\
\vspace{.5cm}
\includegraphics[keepaspectratio=true, width=1\textwidth]{CurriculumStudy/LabInstructionWithRooms.pdf}
\caption{\label{fig:InstSched} Instructor and Room Schedule}
\end{figure}

Assigning one instructor to each curriculum would confound the instructor effect and the curriculum effect.  To avoid confounding, each treatment group would need to receive instruction from both instructors.  An alternating weekly schedule was decided upon to spread out the instructor effects over both curricula. A coin was flipped to decide how to match the instructor to the curriculum when the alternation was initialized. The lecture and lab instruction schedules for each cohort can also be found below in Figure~\ref{fig:InstSched}. Note that each figure has student cohort and times fixed across all weeks, reflecting the unchanged time structure that each student enrolled into. The instructors and room locations are what changed throughout the course.

%-----------------------------------------------------
\subsection{Data Collection}
\label{datacollect}

In order to measure learning outcomes for specific inference concepts we utilized question sets from the Assessment Resource Tools for Improving Statistical Thinking (ARTIST) for the topics of confidence intervals and hypothesis testing \citep{ARTIST}.  The ARTIST scaled question sets each consist of 10 multiple choice questions that are geared toward critical thinking about the inference topic.  These questions were administered as part of the written final exam for all students on the same day and time. The ARTIST scaled scores for the topics of confidence interval and hypothesis testing were recorded for each student. The multiple choice questions for the ARTIST scaled topics can be found in Appendix~\ref{appendA}.

The final exam also included two problems that tested the student's ability to conduct statistical inference in an applied setting using theory-based methodology. Each problem was based on a hypothetical scenario where data has been collected and inference needed to be conducted using the traditional approach using formulas and tables. The first problem provided data summaries and students needed to construct and interpret a confidence interval for a single population mean. The second problem required students to conduct a hypothesis test for a single proportion based on another set of data summaries. The applied inference problem scores for each student are not used for the primary analysis on learning outcomes but are included for an interesting peripheral analysis on student ability to conduct inference using traditional theory-based methods. The applied inference problems and grading rubrics can also be found in Appendix~\ref{appendA}. The exams were graded blindly, with no identifying information of the student or treatment visible during the grading process.

In addition to the ARTIST and applied inference question scores, data were collected from the first eight weeks of the course -- prior to student exposure to an inference curriculum. We have scores from homework assignments 1 to 7, lab assignments 1 to 7 and the midterm exam for each student. The midterm exam questions and grading rubrics can be found in Appendix~\ref{appMidtermExamQuestions}. Since all of these items were administered and graded equivalently for all students before being assigned to a curricula, the scores from these weeks will be referred to as the ``pre-treatment measurements''. Lastly, the data include the cohort to which each student belonged.

The research proposal approved by the Institutional Review Board specified that students' data would be entirely deidentified following the course, including all demographic information.  At the conclusion of the semester the data for the 101 students who consented to the release of their data were saved, with names and identity information removed, to a spreadsheet.  The deidentified student data was imported to \texttt{R} for the analysis described in Section~\ref{analysis} below.

%-----------------------------------------------------
\subsection{Data Summary}
\label{dataSummary}

The ARTIST and applied inference question scores from the final exam are the response variables on which we wish to compare the groups of students from the two inference curricula. Figure~\ref{fig:ScoreHistsByGroup} displays the histogram, mean and standard deviation for each response variable, separated by curricula.  Midterm exam scores as also included in order to provide a comparison of the curricula groups using a pre-treatment measurement.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[H]

{\centering \includegraphics[width=1\linewidth]{FigureScoreHistsByGroup-1}

}

\caption[Histograms and summary statistics of scores separated by curricula group]{Histograms and summary statistics of scores separated by curricula group.}\label{fig:ScoreHistsByGroup}
\end{figure}

\end{knitrout}
\vspace{.05in}
In Figure~\ref{fig:ScoreHistsByGroup}, we see that the midterm exam scores are very similarly distributed for each group; with the traditional curriculum group scoring only slightly higher on average than the simulation-based curriculum group. This similarity is expected -- and desirable -- because the midterm was conducted prior to the treatment being administered, and the class materials and instruction were designed to be identical at that stage of the course.

Comparing the distributions in Figure~\ref{fig:ScoreHistsByGroup} we see that the simulation-based inference group had a higher average score than the traditional inference group on both of the ARTIST scaled question sets and on the applied confidence interval problem, but scored lower on average on the applied hypothesis testing problem.  The simulation-based inference group had lower variability than the traditional inference group on the ARTIST question set for confidence intervals, but higher variability on all other scores. These data summaries are suggestive of differences in the inference learning outcomes of the two groups. In Section~\ref{analysis}, we take a model-based approach to assess if these differences are statistically significant.

% 
% % latex table generated in R 3.0.2 by xtable 1.7-3 package
% % Mon Nov 24 15:50:03 2014
% 
% \begin{table}[ht]
% \small
% \centering
% \begin{tabular}{llrrrrr}
%   \hline
%     & Curriculum & Midterm & ARTIST CI & ARTIST HT & Applied CI & Applied HT \\ 
%   \hline
%   & Traditional & roomdat[1,3] & roomdat[1,4] & roomdat[1,5] & roomdat[1,6] & roomdat[1,7] \\ 
%    & Simulation & roomdat[2,3] & roomdat[2,4] & roomdat[2,5] & roomdat[2,6] & roomdat[2,7] \\ 
%  \hline
%  & & & & & &  \\ 
%  & & & & & &  \\ 
%  \hline
%   
%  Midterm Quartile & Curriculum & Midterm & ARTIST CI & ARTIST HT & Applied CI & Applied HT \\ 
%   \hline
% 1st Quartile & Traditional & 55.50 & 55.38 & 56.92 & 81.41 & 61.54 \\ 
%  & Simulation & 55.68 & 64.29 & 50.00 & 74.40 & 62.34 \\ 
%  & Difference & -0.18 & -8.90 & 6.92 & 7.01 & -0.80 \\ 
%  \hline
% 2nd Quartile & Traditional & 73.19 & 57.69 & 51.54 & 77.56 & 69.23 \\ 
%  & Simulation & 71.00 & 71.67 & 52.50 & 86.11 & 66.67 \\ 
%  & Difference & 2.19 & -13.97 & -0.96 & -8.55 & 2.56 \\ 
%  \hline
% 3rd Quartile & Traditional & 84.58 & 76.15 & 57.69 & 88.46 & 90.91 \\ 
%  & Simulation & 82.71 & 71.67 & 61.67 & 84.72 & 75.00 \\ 
%  & Difference & 1.87 & 4.49 & -3.97 & 3.74 & 15.91 \\ 
%  \hline
% 4th Quartile & Traditional & 94.33 & 78.33 & 65.83 & 90.97 & 93.94 \\ 
%  & Simulation & 91.67 & 82.50 & 76.67 & 95.83 & 93.18 \\ 
%  & Difference & 2.67 & -4.17 & -10.83 & -4.86 & 0.76 \\ 
%    \hline
% \end{tabular}
% \caption{Average percentage scores broken down by midterm exam quartiles.}
% \label{tab:examquartiles}
% \normalsize
% \end{table}

%---------------------------------------------------------------------------
\section{Analysis}
\label{analysis}

The primary goal of the analysis is to investigate if there is a curricula effect on inference concept learning outcomes.  Our data includes ARTIST scaled topic scores for confidence intervals and hypothesis tests which we use as the responses for the comparison of curricula. A model based approach is used to assess curricula effect while controlling for pre-treatment differences between students. With the two dimensional response and an assortment of covariates we employ a multivariate analysis of covariance (MANCOVA) model.

Both curricula groups were required to learn how to conduct normal-based inference. This leads to another question of interest. Does the added simulation-based material turn out to be detrimental to student's ability to use distributional theory-based methods to conduct inference?  Two applied problems were included on the final exam that required students to use theory-based methods and formulas to conduct inference. These applied questions were used as the responses in a separate MANCOVA model to check for a curriculum effect.

The bivariate MANCOVA models used for these two analyses are parameterized as

\begin{eqnarray}\label{eq:mancova}
y_{i\ell} = \tau_{\ell} \mathbbm{1}_{\{i \in T\}} + \beta_{\ell 0} + \sum_{p=1}^{P}x_{ip}\beta_{\ell p} + \epsilon_{i\ell},
\end{eqnarray}
where

\begin{tabular}{lp{5in}}
$y_{i\ell}$ &  is the $\ell^{th}$ response ($\ell \in \left\{1,2\right\}$) from  student $i$, $ 1 \le i \le n$,\\
$\tau_{\ell}$ & is the treatment effect of the simulation-based curriculum on response $\ell$, and \\
$\mathbbm{1}_{\{i \in T\}}$ & is an indicator function, that is one, if student $i$ is in the treatment group, and zero otherwise.\\
$\beta_{\ell 0}$ & is the common intercept for response $\ell$, and \\
$\beta_{\ell p}$, & $1 \le p \le P$ are the model coefficients of the $P$ covariates. \\
$x_{ip}$ & is the $p^{th}$ pre-treatment covariate score of student $i$, and\\
$\epsilon}_{i\ell}$ & is the error for the $\ell ^{th}$ response from the $i^{th}$ student. 
\end{tabular}

We assume that error pairs are independent and identically distributed:
\[
\vec{\epsilon}_{i} = 
\begin{bmatrix}
  \epsilon_{i1} \\ \epsilon_{i2} 
 \end{bmatrix}  
 \distas{iid} \text{MVN} \left( 
 \begin{bmatrix}
  0 \\ 0 
 \end{bmatrix},
 \Sigma = \begin{bmatrix}
  \sigma_{11}^2 & \sigma_{12}^2 \\ 
  \sigma_{21}^2 & \sigma_{22}^2
 \end{bmatrix}
 \right)
\]

% \begin{center}
% $y_{ik} = \tau_{k} \mathbbm{1}_{\{i \in T\}} + \beta_{0k} + \Bigg(\sum_{p=1}^{P}x_{ip}\beta_{pk}\Bigg) + \epsilon_{ik}$,\\
% 
% where $k \in \left\{1,2\right\}$,  $i \in \left\{1,...,n\right\}$, and $p \in \left\{1,...,P\right\}$,
% 
% $\vec{\epsilon}_{i} = 
% \begin{bmatrix}
%   \epsilon_{i1} \\ \epsilon_{i2} 
%  \end{bmatrix}  
%  \distas{iid} $ MVN$\left( 
%  \begin{bmatrix}
%   0 \\ 0 
%  \end{bmatrix},
%   \begin{bmatrix}
%   \sigma_{11}^2 & \sigma_{12}^2 \\ 
%   \sigma_{21}^2 & \sigma_{22}^2
%  \end{bmatrix}
%  \right)$,
% \end{center}

With this parameterization, it is clear that the underlying structure of the MANCOVA model is a multivariate multiple linear regression that can include categorical and continuous covariates. Note that the paired error terms from each student are correlated but are specified as independent between students. The assumption of independence between student response scores is understood to be unrealistic for students from the same class. \km{The repercussions of violating the assumption of independence between student responses will be explored through a simulation study in Section~\ref{SimStudy} following the analysis.}

%-----------------------------------------------------
\subsection{Modeling ARTIST Outcomes}
\label{ArtistModel}

We begin with the model for the ARTIST scaled topic scores.  Many of the pre-treatment variables are highly correlated. To select a model with only the most predictive pre-treatment covariates, model selection was conducted by first running backward selection based on AIC then removing further covariates that posed collinearity issues. The model selected for final analysis included three covariates: an indicator variable for the curriculum treatment group, the lab 5 score and the midterm score.  The midterm tested students on materials from weeks 1-7 and lab 5 assessed understanding of topics related to random selection techniques.  We will refer to this selected model as the ``ARTIST Model''. Model fit for the ARTIST Model was assessed to be satisfactory; see Appendix~\ref{appARTISTModDiag} for residual plots and other model diagnostics.

To test for overall covariate effects on the multivariate responses we use Pillai's $\Lambda$. Table~\ref{tab:overallmod} shows a weak overall effect of the curriculum treatment on the paired ARTIST scaled topic scores.  This prompts us to investigate the treatment effect on the ARTIST scaled topic scores for confidence intervals and hypothesis tests separately, to see if the weak overall effect is driven by a significant effect on one of the two scores.

\begin{table}[hbtp]
\centering
\begin{tabular}{lrrr} \hline
 & Pillai's $\Lambda$ & Approx. F & Pr($>$$|$F$|$)\\ 
 \hline  
Midterm & 0.2109 & 12.8277 &
0.0000\\ 
Lab 5 & 0.0792 & 4.1261 &
0.0191\\ 
Treatment & 0.0469 & 2.3605 &
0.0998\\ 
\hline
\end{tabular}
\caption{Tests for overall covariate effects on ARTIST question scores using Pillai's $\Lambda$.}
\label{tab:overallmod}
\end{table}

To investigate the effect of the curriculum treatment on each ARTIST scaled topic score, we analyze the two underlying linear models that comprise the overall MANCOVA model. Table~\ref{tab:cimod} displays the coefficients of the linear model fit to the ARTIST scaled score for confidence interval learning outcomes along with covariate ranges to provide context to coefficient magnitudes. It should be noted that although the midterm and lab scores were recorded discretely, they were treated as continuous covariates when fitting the model. We find that midterm, lab 5 and the curriculum treatment effect are significant. Specifically, the simulation-based inference group scored significantly higher by 0.7149 out of a possible 10 points, a 7.146\% improvement in confidence interval learning outcomes on the ARTIST scale while controlling for midterm and lab 5 scores.\\

% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Tue Jul 29 15:46:21 2014
\begin{table}[hbtp]
\centering
\begin{tabular}{rccc}
  \hline
 & Covariate Values & Estimate & 95\% Confidence Interval \\ 
  \hline
Intercept & 1 & 1.4648 & ( -0.5467 , 3.4763 ) \\ 
  Midterm & \{0,1,...100\} & 0.0477 & (  0.0253 , 0.0701 ) \\ 
  Lab 5  &  \{0,1,...100\} & 0.0183 & (  0.0057 , 0.0309 ) \\ 
  Treatment &  \{0,1\} & 0.7146 & (  0.0435 , 1.3858 ) \\ 
   \hline
\end{tabular}
\caption{Coefficients for model fit to ARTIST score for confidence interval topic.}
\label{tab:cimod}
\end{table}

Table~\ref{tab:htmod} displays the coefficients of the linear model fit to the ARTIST scaled score for hypothesis test learning outcomes.  We find that only the midterm score is significant for predicting learning outcomes for hypothesis testing. There was no significant curriculum treatment effect. \\

% latex table generated in R 3.0.2 by xtable 1.7-3 package
% Tue Jul 29 15:41:44 2014
\begin{table}[hbtp]
\centering
\begin{tabular}{rccc}
  \hline
 & Covariate Values & Estimate & 95\% Confidence Interval \\ 
  \hline
Intercept & 1 & 2.1053 & ( 0.0046 , 4.2060 ) \\ 
  Midterm & \{0,1,...100\} & 0.0386 & (  0.0152 , 0.0620 ) \\ 
  Lab 5  &  \{0,1,...100\} & 0.0085 & (  -0.0046 , 0.0217 ) \\ 
  Treatment &  \{0,1\} & 0.3050 & (  -0.3960 , 1.0059 ) \\ 
   \hline
\end{tabular}
\caption{Coefficients for model fit to ARTIST score for hypothesis test topic.}
\label{tab:htmod}
\end{table}
